High-level cognition during story listening is reflected in high-order dynamic correlations in neural activity patterns

### 2021-01-15 paper/main.tex
added requested refs to intro and discussion, updated main and diffs

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features (Fig.~\ref{fig:patterns}a) can also help to elucidate which
specific aspects of neural activity patterns are informative about
cognition and, by extension, which types of neural activity patterns
might compose the neural code.  For example, prior work has used
region of interest analyses to estimate the anatomical locations of
specific neural representations~\citep[e.g.,][]{EtzeEtal09}, or to
compare the relative contributions to the neural code of multivariate
activity patterns versus dynamic correlations between neural activity
patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An emerging theme
in this literature is that cognition is mediated by dynamic
interactions between brain structures~\citep{Gros88, Fris00, SporHone06, BassEtal06,
  Turk13, DemeEtal19, SoloEtal19, LuriEtal18, PretEtal17, ZouEtal19,
  MackEtal17}.

@@

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, DemeEtal19}.  What might the future of this line of work
hold?  For example, is the neural code implemented through
higher-order interactions between brain structures~\citep[e.g.,
see][]{ReimEtal17}?  Second-order correlations reflect
\textit{homologous} patterns of correlation.  In other words, if the
dynamic patterns of correlations between two regions, $A$ and $B$, are
similar to those between two other regions, $C$ and $D$, this would be
reflected in the second-order correlations between ($A$--$B$) and
($C$--$D$).  In this way, second-order correlations identify
similarities and differences between subgraphs of the brain's
connectome.  Analogously, third-order correlations reflect homologies
between second-order correlations-- i.e., homologous patterns of
homologous interactions between brain regions.  More generally,
higher-order correlations reflect homologies between patterns of
lower-order correlations.  We can then ask: which ``orders'' of
interaction are most reflective of high-level cognitive processes?

@@

Another central question pertains to the extent to which the neural
code is carried by activity patterns that directly reflect ongoing
cognition~\citep[e.g., following][]{HaxbEtal01, NormEtal06}, versus
the dynamic properties of the network structure itself, independent of
specific activity patterns in any given set of regions~\citep[e.g.,
following][]{BassEtal06}.  For example, graph measures such as
centrality and degree~\citep{BullSpor09} may be used to estimate how a
given brain structure is ``communicating'' with other structures,
independently of the specific neural representations carried by those
structures.  If one considers a brain region's position in the network
(e.g., its eigenvector centrality) as a dynamic property, one can
compare how the positions of different regions are correlated, and/or
how those patterns of correlations change over time.  We can also
compute higher-order patterns in these correlations to characterize
homologous subgraphs in the connectome that display similar changes in
their constituent brain structures' interactions with the rest of the
brain.

@@

The notion that cognition is reflected in (and possibly mediated by)
patterns of first-order network dynamics has been suggested by or
proposed in myriad empirical studies and
reviews~\citep[e.g.,][]{DemeEtal19, Turk13, LuriEtal18, FongEtal19,
  ParkEtal18b, PretEtal17, MannEtal18, RoyEtal19, LiegEtal19, ZouEtal19,
  ChanGlov10, GonzEtal19}.  Our study extends this line of work by
finding cognitively relevant \textit{higher-order} network dynamics
that reflect ongoing cognition.  Our findings also complement other work
that uses graph theory and topology to characterize how brain networks
reconfigure during cognition~\citep[e.g.,][]{BassEtal06, ZhenEtal19,
  McInJirs19, TokeSomm19, SizeEtal18, ReimEtal17, BetzEtal19}.

An open question not addressed by our study pertains to how different
structures integrate incoming information with different time
constants.  For example, one line of work suggests that the cortical
surface comprises a structured map such that nearby brain structures
process incoming information at similar timescales.  Low-level sensory
areas integrate information relatively quickly, whereas higher-level
regions integrate information relatively slowly~\citep{BaldEtal17,
  HassEtal08, HassEtal15, HoneEtal12a, LernEtal11, LernEtal14,
  ChieHone19}.  A similar hierarchy appears to play a role in
predicting future events~\citep{LeeEtal20}.  Other related work in
human and mouse brains indicates that the temporal response profile of
a given brain structure may relate to how strongly connected that
structure is with other brain areas~\citep{FallEtal19}.  Further study
is needed to understand the role of temporal integration at different
scales of network interaction, and across different anatomical
structures. Importantly, our analyses do not speak to the
physiological basis of higher-order dynamics, and could reflect
nonlinearities, chaotic patterns, non-stationarities, and/or
multistability, etc. However, our decoding analyses do indicate that
higher-order dynamics are consistent across individuals, and therefore
unlikely to reflect non-stimulus-driven dynamics that are unlikely to
be similar across individuals.

@@

The modern toolkit of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA;][]{Pear01}, Probabilistic
PCA~\citep[PPCA;][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA;][]{Spea04}, Independent Components
Analysis~\citep[ICA;][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE;][]{MaatHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP;][]{McInEtal18}, non-negative
matrix factorization~\citep[NMF;][]{LeeSeun99}, Topographic Factor
Analysis~\cite[TFA;][]{MannEtal14b}, Hierarchical Topographic Factor
analysis~\cite[HTFA;][]{MannEtal18}, Topographic Latent Source
Analysis~\cite[TLSA;][]{GersEtal11}, dictionary
learning~\citep{MairEtal09a, MairEtal09b}, and deep
auto-encoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $K$ matrix, $\mathbf{X}$, that is closest to
the original $T$ by $J$ matrix, $\mathbf{Y}$, where (typically)
$K \ll J$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of the
data are compared (and how) in order to optimize how well
$\mathbf{X}$ approximates $\mathbf{Y}$.

@@

\paragraph{Hierarchical topographic factor analysis (HTFA).}
Following our prior related work, we used HTFA~\citep{MannEtal18} to
derive a compact representation of the neuroimaging data.  In brief,
this approach approximates the timeseries of voxel activations (44,415
voxels) using a much smaller number of radial basis function (RBF)
nodes~\citep[in this case, 700 nodes, as determined by an optimization
procedure described by][]{MannEtal18}.  This provides a convenient
representation for examining full-brain network dynamics.  All of the
analyses we carried out on the neuroimaging dataset were performed in
this lower-dimensional space.  In other words, each participant's data
matrix, $\mathbf{X}_0$, was a number-of-timepoints by 700 matrix of
HTFA-derived factor weights (where the row and column labels were
matched across participants).  Code for carrying out HTFA on fMRI data
may be found as part of the BrainIAK toolbox~\citep{brainiak}, which
may be downloaded at \href{https://brainiak.org/}{brainiak.org}.

@@

\bibliographystyle{apacite}
\bibliography{CDL-bibliography/memlab}

>>>>

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06b, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features (Fig.~\ref{fig:patterns}a) can also help to elucidate which
specific aspects of neural activity patterns are informative about
cognition and, by extension, which types of neural activity patterns
might compose the neural code.  For example, prior work has used
region of interest analyses to estimate the anatomical locations of
specific neural representations~\citep[e.g.,][]{EtzeEtal09}, or to
compare the relative contributions to the neural code of multivariate
activity patterns versus dynamic correlations between neural activity
patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An emerging theme
in this literature is that cognition is mediated by dynamic
interactions between brain structures~\citep{Gros88, Fris00, SporHone06, BassEtal06,
  Turk13, DemeEtal19, SoloEtal19, LuriEtal18, PretEtal17, ZouEtal19,
  MackEtal17, BresKels01, McIn00}.

@@

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\citep[for review
see][]{NormEtal06b}, or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19, LuriEtal18, PretEtal17,
  ZouEtal19, DemeEtal19}.  What might the future of this line of work
hold?  For example, is the neural code implemented through
higher-order interactions between brain structures~\citep[e.g.,
see][]{ReimEtal17}?  Second-order correlations reflect
\textit{homologous} patterns of correlation.  In other words, if the
dynamic patterns of correlations between two regions, $A$ and $B$, are
similar to those between two other regions, $C$ and $D$, this would be
reflected in the second-order correlations between ($A$--$B$) and
($C$--$D$).  In this way, second-order correlations identify
similarities and differences between subgraphs of the brain's
connectome.  Analogously, third-order correlations reflect homologies
between second-order correlations-- i.e., homologous patterns of
homologous interactions between brain regions.  More generally,
higher-order correlations reflect homologies between patterns of
lower-order correlations.  We can then ask: which ``orders'' of
interaction are most reflective of high-level cognitive processes?

@@

Another central question pertains to the extent to which the neural
code is carried by activity patterns that directly reflect ongoing
cognition~\citep[e.g., following][]{HaxbEtal01, NormEtal06b}, versus
the dynamic properties of the network structure itself, independent of
specific activity patterns in any given set of regions~\citep[e.g.,
following][]{BassEtal06}.  For example, graph measures such as
centrality and degree~\citep{BullSpor09} may be used to estimate how a
given brain structure is ``communicating'' with other structures,
independently of the specific neural representations carried by those
structures.  If one considers a brain region's position in the network
(e.g., its eigenvector centrality) as a dynamic property, one can
compare how the positions of different regions are correlated, and/or
how those patterns of correlations change over time.  We can also
compute higher-order patterns in these correlations to characterize
homologous subgraphs in the connectome that display similar changes in
their constituent brain structures' interactions with the rest of the
brain.

@@

The notion that cognition is reflected in (and possibly mediated by)
patterns of first-order network dynamics has been suggested by or
proposed in myriad empirical studies and
reviews~\citep[e.g.,][]{DemeEtal19, Turk13, LuriEtal18, FongEtal19,
  ParkEtal18b, PretEtal17, MannEtal18, RoyEtal19, LiegEtal19, ZouEtal19,
  ChanGlov10, GonzEtal19, McIn00, BresKels01}.  Our study extends this line of work by
finding cognitively relevant \textit{higher-order} network dynamics
that reflect ongoing cognition.  Our findings also complement other work
that uses graph theory and topology to characterize how brain networks
reconfigure during cognition~\citep[e.g.,][]{BassEtal06, ZhenEtal19,
  McInJirs19, TokeSomm19, SizeEtal18, ReimEtal17, BetzEtal19}.

An open question not addressed by our study pertains to how different
structures integrate incoming information with different time
constants.  For example, one line of work suggests that the cortical
surface comprises a structured map such that nearby brain structures
process incoming information at similar timescales.  Low-level sensory
areas integrate information relatively quickly, whereas higher-level
regions integrate information relatively slowly~\citep{BaldEtal17,
  HassEtal08, HassEtal15, HoneEtal12a, LernEtal11, LernEtal14,
  ChieHone19}.  A similar hierarchy appears to play a role in
predicting future events~\citep{LeeEtal20}.  Other related work in
human and mouse brains indicates that the temporal response profile of
a given brain structure may relate to how strongly connected that
structure is with other brain areas~\citep{FallEtal20}.  Further study
is needed to understand the role of temporal integration at different
scales of network interaction, and across different anatomical
structures. Importantly, our analyses do not speak to the
physiological basis of higher-order dynamics, and could reflect
nonlinearities, chaotic patterns, non-stationarities, and/or
multistability, etc. However, our decoding analyses do indicate that
higher-order dynamics are consistent across individuals, and therefore
unlikely to reflect non-stimulus-driven dynamics that are unlikely to
be similar across individuals.

@@

The modern toolkit of dimensionality reduction algorithms include
Principal Components Analysis~\citep[PCA;][]{Pear01}, Probabilistic
PCA~\citep[PPCA;][]{TippBish99}, Exploratory Factor
Analysis~\citep[EFA;][]{Spea04}, Independent Components
Analysis~\citep[ICA;][]{JuttHera91, ComoEtal91}, $t$-Stochastic
Neighbor Embedding~\citep[$t$-SNE;][]{vandHint08}, Uniform Manifold
Approximation and Projection~\citep[UMAP;][]{McInEtal18}, non-negative
matrix factorization~\citep[NMF;][]{LeeSeun99}, Topographic Factor
Analysis~\cite[TFA;][]{MannEtal14b}, Hierarchical Topographic Factor
analysis~\cite[HTFA;][]{MannEtal18}, Topographic Latent Source
Analysis~\cite[TLSA;][]{GersEtal11}, dictionary
learning~\citep{MairEtal09a, MairEtal09b}, and deep
auto-encoders~\citep{HintSala06}, among others.  While complete
characterizations of each of these algorithms is beyond the scope of
the present manuscript, the general intuition driving these approaches
is to compute the $T$ by $K$ matrix, $\mathbf{X}$, that is closest to
the original $T$ by $J$ matrix, $\mathbf{Y}$, where (typically)
$K \ll J$.  The different approaches place different constraints on
what properties $\mathbf{X}$ must satisfy and which aspects of the
data are compared (and how) in order to optimize how well
$\mathbf{X}$ approximates $\mathbf{Y}$.

@@

\paragraph{Hierarchical topographic factor analysis (HTFA).}
Following our prior related work, we used HTFA~\citep{MannEtal18} to
derive a compact representation of the neuroimaging data.  In brief,
this approach approximates the timeseries of voxel activations (44,415
voxels) using a much smaller number of radial basis function (RBF)
nodes~\citep[in this case, 700 nodes, as determined by an optimization
procedure described by][]{MannEtal18}.  This provides a convenient
representation for examining full-brain network dynamics.  All of the
analyses we carried out on the neuroimaging dataset were performed in
this lower-dimensional space.  In other words, each participant's data
matrix, $\mathbf{X}_0$, was a number-of-timepoints by 700 matrix of
HTFA-derived factor weights (where the row and column labels were
matched across participants).  Code for carrying out HTFA on fMRI data
may be found as part of the BrainIAK toolbox~\citep{CapoEtal17}, which
may be downloaded at \href{https://brainiak.org/}{brainiak.org}.

@@

\bibliographystyle{apacite}
\bibliography{CDL-bibliography/cdl}
